### 2023-02-09 definitions.tex
details

%
%\problemx{Bounded Resilience problem}
%{a transition system $(S,\rightarrow)$, a state $s \in S$, $SAFE, BAD \subseteq S$ and $SAFE \cap BAD = \emptyset$.}
%{$\exists k \geq 0 ~ \forall s' \in BAD, ~ s \rightarrow^* s' \implies \exists s'' \in SAFE ~ s' \rightarrow^{\leq k} s''$ ?\newline}
%

\problemx{Bounded Resilience problem}
{a transition system $(S,\rightarrow)$, a state $s \in S$, $SAFE, BAD \subseteq S$ and $SAFE \cap BAD = \emptyset$.}
{$\exists k \geq 0  \mathscr{S}$ is $k$-resilient. ?\newline}
%%
\fi
%%%

pour $k,k'$ donnés, chercher si $SAFE_{max}$ et $BAD_{max}$ ont un sens ? on peut faire l'union des SAFE, SAFE' et des BAD, BAD' prendre le max des k. Fixer k, chercher et calculer $SAFE_{max}$ et $BAD_{max}$.

%
%\alain{dire juste que SAFE = upward closed et BAD = downward closed.
%envisager les 3 autres cas. souvent BAD = upward closed (par ex l'exclusion mutuelle).
%4 resilience problem (U,D), (U,U), (D,U), (D,D)}

@@

The resilience problem (resp. the $k$-resilience problem) in a transition system is to decide whether from a bad state, there exists a path (resp. a path of length smaller than $k$) that reaches a safe state. Let us formalize these properties.

\problemx{uniform resilience problem (URP)}
{A transition system $\mathscr{S}=(S,\rightarrow)$ and two sets $SAFE, BAD \subseteq S$.}
{$BAD \longrightarrow^{*} SAFE$ ?\newline}
%
%\alain{il faudrait ne pas répéter 3 fois les mêmes imputs pour les 3 pbs: énoncer les 3 uniformes pbs d'un coup avec une fois l'input puis les 3 pbs pour un état s donné sans répéter non plus 3 fois les mêmes inputs}

\problemx{uniform $k$-resilience problem (UkRP)}
{A transition system $\mathscr{S}=(S,\rightarrow), k \in \mathbb{N}$ and two sets $SAFE, BAD \subseteq S$.}
{$BAD \longrightarrow^{\leq k} SAFE$ ?\newline}

We add a third problem that decides whether there exists an $k$ such  that the system is $k$-resilient.
%
\problemx{uniform bounded resilience problem (UBRP)}
{A transition system $\mathscr{S}=(S,\rightarrow)$ and two sets $SAFE, BAD \subseteq S$.}
%{$\exists k \geq 0 ~ \forall s' \in D ~ s \rightarrow^* s' \implies \exists s'' \in U ~ s' \rightarrow^{\leq k} s''$ ?\newline}
{$\exists k \geq 0$ such that $\mathscr{S}$ is uniformely $k$-resilient ?\newline}

\alain{trouver les bad et safe maximum tels que S est resilient. est-ce vrai que si S est $(B_i,D_i)$-resilient alors S est $(\cap, \cup B_i,D_i)$-resilient ? on pourrait décider que safe=complement de bad et donc on aurait un seul ensemble (et son complement)}

These three reachability problems are decidable for finite transition systems but undecidable for (general) infinite-state transition systems. 
So we restrict our framework to the class of infinite-state WSTS. Since most of decidable properties in WSTS are given as effective upward or downward closed sets \cite{DBLP:journals/iandc/AbdullaCJT00, DBLP:journals/tcs/FinkelS01}, we consider upward closed or downward closed sets SAFE and BAD.
\alain{on peut penser à des ensembles bad definis dans une logique booleennne sur les clos par le bas, haut, +...}

For instance, the well-known mutual exclusion property is often modelized in a $d$-counters machine by the property that a counter $c_{mutex}$ must be bounded by (usually) one. Then, the set $SAFE =  \{c_{mutex} \leq 1\} \times \mathbb{N}^{d-1}$ is downward closed and $BAD =\{c_{mutex} \geq 2\} \times  \mathbb{N}^{d-1} $ is the upward closed complementary of SAFE. In \cite{DBLP:conf/gg/Ozkan22}, the authors considered that BAD is always downward closed and SAFE is always upward closed.
%		
URP is decidable for lossy counter machines (LCM) with SAFE and BAD semilinear sets as a consequence of results in Section 3.4 of \cite{DBLP:conf/rp/Schnoebelen10}. In particular, the existing proofs of resilience use the decidability of reachability; but we wish to decide resilience for models with undecidable reachability.
%
In what follows, we will always suppose that a downward closed set $D$ is given by its finite set of ideals $I_D=\{J_1, J_2,...,J_n\}$ satisfying $D=\downarrow D = \downarrow (J_1 \cup J_2 \cup..\cup J_n)$ and that an upward closed set $U$ is given by its finite set $Min_U$ of its minimal elements satisfying $U=\uparrow Min_U$.

@@

\begin{proposition}\label{general}
$\mathscr{S}=(S,\rightarrow,\leq)$ is uniformely (bad,safe)-resilient iff $bad \subseteq \pred^*(safe)$.\\
$\mathscr{S}=(S,\rightarrow,\leq)$ is uniformely (bad,safe)-bounded-resilient iff $\mathscr{S}=(S,\rightarrow,\leq)$ is uniformely (bad,safe)-resilient.\\
$\mathscr{S}=(S,\rightarrow,\leq)$ is uniformely (bad,safe)-$k$-resilient iff $bad \subseteq \pred^k(safe)$.
\end{proposition}

@@

\subsection{Case: $SAFE=\uparrow SAFE$.}

\begin{theorem}\label{up-down}
The three uniform resilience problems are decidable for WSTS with effective predbasis and $safe=\uparrow safe$.
%	and bad is upward closed or downward closed.
\end{theorem}

\begin{proof}
Let us first solve if $\mathscr{S}=(S,\rightarrow,\leq)$ is uniformely (bad,safe)-resilient. Since $safe=\uparrow safe$, and WSTS, $\pred^*(safe)=\uparrow \pred^*(safe)$, so $\pred^*(safe)$ admits a finite basis $B_{\pred^*(safe)}$. since WSTS  with effective predbasis, we may compute a finite basis of $\pred^*(safe)$ with the backward coverability algorithm. 
%Then if bad is upward closed or downward closed, we may decide the inclusion $Bad \subseteq \uparrow B_{\pred^*(safe)}$ which remais to compare, in both cases, two finite sets. 
if bad  is upward closed, $bad = \uparrow B_{bad}$ and $ \uparrow B_{bad} \subseteq \uparrow B_{\pred^*(safe)}$ iff for every $b \in B_{bad}$, there is a $s \in B_{\pred^*(safe)}$ such that $s \leq b$. similar reasonning for the other case.
hence the two uniform resilience problems are decidable.

Let us examine the uniform $k$-resilience problem (UkRP). we may compute the least $n$ such that $bad \subseteq \uparrow \pred^n(SAFE)=  \uparrow \pred^*(SAFE)$. S is uniformely $k$-resilient iff $n \leq k$ \alain{hum...mais il peut y avoir des chemins plus courts}and then UkRP is decidable.
\end{proof}

%One begins to test whether $Bad \cap \pred^*(safe) = \emptyset$ then never resilient car une fois dans bad j'y reste sans pouvoir aller dans safe.\\

If $Bad \cap \pred^*(safe) \neq \emptyset$, One computes the minimal elements $A$ of $Bad \cap \pred^*(safe)$.
resilient iff $Bad \subseteq \pred^*(safe)$  (eq $Bad \cap \pred^*(safe) = Bad$)

$Bad \cap \pred^*(safe) \neq Bad$ alors, on peut se limiter à $Bad' = Bad \cap \pred^*(safe)$ et c'est (Bad',safe)-resilient.

@@

\subsection{Case: $SAFE=\uparrow SAFE$ and $BAD=\downarrow BAD$.}

@@

%Transfering the abstract resilience problems into this framework,
%it is therefore reasonable to demand that both propositions, SAFE and BAD, are given by 
%upward-closed or downward-closed sets.

@@

\begin{proposition}\label{down-up}
Let $\mathscr{S}=(S,\rightarrow, \leq)$ be a completion-post-effective $\omega^2$-WSTS with strong compatibility and two finite sets: $I_{BAD}$ and $Min_{SAFE}$.
The uniform resilience problem (URP) and the uniform bounded resilience problem (UBRP) are decidable.
The uniform $k$-resilience problem (UkRP) is decidable if moreover $\mathscr{S}$ has the predbasis hypothesis.
\end{proposition}

\begin{proof}
Let $I_{BAD}=\{J_1, J_2,...,J_n\}$ be the ideal decomposition of $BAD$ and $B_{SAFE}=\{b_1,b_2,...,b_m\}$ be the minimal basis of $SAFE$.
The uniform resilience problem (URP) is equivalent to an infinite number of instances of the coverability problem : for all $x \in BAD$ does there exist an $j$ such that $b_j$ is coverable from $x$. This infinite set of coverability questions can be reduced to $n$ instances of the coverability problem in the completion $\hat{\mathscr{S}}=(Ideals(S),\rightarrow, \subseteq)$ of $\mathscr{S}=(S,\rightarrow, \leq)$.

@@

To decide the UkRP, we may compare the input $k$ with $K$. If $k \geq K$, we conclude that is uniformely $k$-resilient; if $k < K$ we cannot answer. if moreover $\mathscr{S}=(S,\rightarrow, \leq)$ has the predbasis hypothesis, we may compute the least $n$ such that $ \uparrow \pred^n(SAFE)=  \uparrow \pred^*(SAFE)$. S is uniformely $k$-resilient iff $n \leq k$ \alain{hum...mais il peut y avoir des chemins plus courts}and then UkRP is decidable.
\end{proof}

@@

\subsection{Case: $SAFE=\uparrow SAFE$ and $BAD=\uparrow BAD$.}

@@

\subsection{Case: $SAFE=\downarrow SAFE$ and $BAD=\downarrow BAD$.}

%%%%%%%%
\subsection{Case: $SAFE=\downarrow SAFE$ and $BAD=\uparrow BAD$.}
%
%		cas exclusion mutuelle
%
je prends les minimaux de bad et les idéaux de safe, pas sur que ce soit decidable....$pred(safe)$

\begin{tabular}{ l c r }
   BAD/SAFE & Up & Down \\
   Up & dec ? & dec ? \\
   Down & dec & ? \\
 \end{tabular}

@@

%
%\problemx{(I,J)-$k$-resilience problem for WSTS}
%{A state $s$ of a WSTS $(S,\rightarrow, \leq)$, an effective set $I$ (with a given basis), a set $J$.}
%{$\forall s' \in J ~ (s \rightarrow^* s') \implies \exists s'' \in I ~ s' \rightarrow^{\leq k} s''$ ?\newline}
%
%\problemx{general $k$-resilience problem for WSTS}
%{A state $s$ of a WSTS $(S,\rightarrow, \leq)$, an effective set $I$ (with a given basis).}
%{$\forall s'  ~ (s \rightarrow^* s') \implies \exists s'' \in I ~ s' \rightarrow^{\leq k} s''$ ?\newline}

>>>>

%
%\problemx{Bounded Resilience problem}
%{a transition system $(S,\rightarrow)$, a state $s \in S$, $Safe, BAD \subseteq S$ and $Safe \cap BAD = \emptyset$.}
%{$\exists k \geq 0 ~ \forall s' \in BAD, ~ s \rightarrow^* s' \implies \exists s'' \in Safe ~ s' \rightarrow^{\leq k} s''$ ?\newline}
%

\problemx{Bounded Resilience problem}
{a transition system $(S,\rightarrow)$, a state $s \in S$, $Safe, Bad \subseteq S$ and $Safe \cap Bad = \emptyset$.}
{$\exists k \geq 0  \mathscr{S}$ is $k$-resilient. ?\newline}
%%
\fi
%%%

pour $k,k'$ donnés, chercher si $Safe_{max}$ et $Bad_{max}$ ont un sens ? on peut faire l'union des Safe, Safe' et des Bad, Bad' prendre le max des k. Fixer k, chercher et calculer $Safe_{max}$ et $Bad_{max}$.

%
%\alain{dire juste que Safe = upward closed et BAD = downward closed.
%envisager les 3 autres cas. souvent BAD = upward closed (par ex l'exclusion mutuelle).
%4 resilience problem (U,D), (U,U), (D,U), (D,D)}

@@

The resilience problem (resp. the $k$-resilience problem) in a transition system is to decide whether from a bad state, there exists a path (resp. a path of length smaller than $k$) that reaches a Safe state. Let us formalize these properties.

\problemx{uniform resilience problem (URP)}
{A transition system $\mathscr{S}=(S,\rightarrow)$ and two sets $Safe, Bad \subseteq S$.}
{$Bad \longrightarrow^{*} Safe$ ?\newline}
%
%\alain{il faudrait ne pas répéter 3 fois les mêmes imputs pour les 3 pbs: énoncer les 3 uniformes pbs d'un coup avec une fois l'input puis les 3 pbs pour un état s donné sans répéter non plus 3 fois les mêmes inputs}

\problemx{uniform $k$-resilience problem (UkRP)}
{A transition system $\mathscr{S}=(S,\rightarrow), k \in \mathbb{N}$ and two sets $Safe, Bad \subseteq S$.}
{$Bad \longrightarrow^{\leq k} Safe$ ?\newline}

We add a third problem that decides whether there exists an $k$ such  that the system is $k$-resilient.
%
\problemx{uniform bounded resilience problem (UBRP)}
{A transition system $\mathscr{S}=(S,\rightarrow)$ and two sets $Safe, Bad \subseteq S$.}
%{$\exists k \geq 0 ~ \forall s' \in D ~ s \rightarrow^* s' \implies \exists s'' \in U ~ s' \rightarrow^{\leq k} s''$ ?\newline}
{$\exists k \geq 0$ such that $\mathscr{S}$ is uniformely $k$-resilient ?\newline}

\alain{trouver les Bad et Safe maximum tels que S est resilient. est-ce vrai que si S est $(B_i,D_i)$-resilient alors S est $(\cap, \cup B_i,D_i)$-resilient ? on pourrait décider que Safe=complement de Bad et donc on aurait un seul ensemble (et son complement)}

These three reachability problems are decidable for finite transition systems but undecidable for (general) infinite-state transition systems. 
So we restrict our framework to the class of infinite-state WSTS. Since most of decidable properties in WSTS are given as effective upward or downward closed sets \cite{DBLP:journals/iandc/AbdullaCJT00, DBLP:journals/tcs/FinkelS01}, we consider upward closed or downward closed sets Safe and Bad.
\alain{on peut penser à des ensembles Bad definis dans une logique booleennne sur les clos par le bas, haut, +...}

For instance, the well-known mutual exclusion property is often modelized in a $d$-counters machine by the property that a counter $c_{mutex}$ must be bounded by (usually) one. Then, the set $Safe =  \{c_{mutex} \leq 1\} \times \mathbb{N}^{d-1}$ is downward closed and $Bad =\{c_{mutex} \geq 2\} \times  \mathbb{N}^{d-1} $ is the upward closed complementary of Safe. In \cite{DBLP:conf/gg/Ozkan22}, the authors considered that Bad is always downward closed and Safe is always upward closed.
%		
URP is decidable for lossy counter machines (LCM) with Safe and Bad semilinear sets as a consequence of results in Section 3.4 of \cite{DBLP:conf/rp/Schnoebelen10}. In particular, the existing proofs of resilience use the decidability of reachability; but we wish to decide resilience for models with undecidable reachability.
%
In what follows, we will always suppose that a downward closed set $D$ is given by its finite set of ideals $I_D=\{J_1, J_2,...,J_n\}$ satisfying $D=\downarrow D = \downarrow (J_1 \cup J_2 \cup..\cup J_n)$ and that an upward closed set $U$ is given by its finite set $Min_U$ of its minimal elements satisfying $U=\uparrow Min_U$.

@@

\begin{proposition}\label{general}
$\mathscr{S}=(S,\rightarrow,\leq)$ is uniformely (bad,Safe)-resilient iff $bad \subseteq \pred^*(Safe)$.\\
$\mathscr{S}=(S,\rightarrow,\leq)$ is uniformely (bad,Safe)-bounded-resilient iff $\mathscr{S}=(S,\rightarrow,\leq)$ is uniformely (bad,Safe)-resilient.\\
$\mathscr{S}=(S,\rightarrow,\leq)$ is uniformely (bad,Safe)-$k$-resilient iff $bad \subseteq \pred^k(Safe)$.
\end{proposition}

@@

\subsection{Case: $Safe=\uparrow Safe$.}

\begin{theorem}\label{up-down}
The three uniform resilience problems are decidable for WSTS with effective predbasis and $Safe=\uparrow Safe$.
%	and bad is upward closed or downward closed.
\end{theorem}

\begin{proof}
Let us first solve if $\mathscr{S}=(S,\rightarrow,\leq)$ is uniformely (bad,Safe)-resilient. Since $Safe=\uparrow Safe$, and WSTS, $\pred^*(Safe)=\uparrow \pred^*(Safe)$, so $\pred^*(Safe)$ admits a finite basis $B_{\pred^*(Safe)}$. since WSTS  with effective predbasis, we may compute a finite basis of $\pred^*(Safe)$ with the backward coverability algorithm. 
%Then if bad is upward closed or downward closed, we may decide the inclusion $Bad \subseteq \uparrow B_{\pred^*(Safe)}$ which remais to compare, in both cases, two finite sets. 
if bad  is upward closed, $bad = \uparrow B_{bad}$ and $ \uparrow B_{bad} \subseteq \uparrow B_{\pred^*(Safe)}$ iff for every $b \in B_{bad}$, there is a $s \in B_{\pred^*(Safe)}$ such that $s \leq b$. similar reasonning for the other case.
hence the two uniform resilience problems are decidable.

Let us examine the uniform $k$-resilience problem (UkRP). we may compute the least $n$ such that $bad \subseteq \uparrow \pred^n(Safe)=  \uparrow \pred^*(Safe)$. S is uniformely $k$-resilient iff $n \leq k$ \alain{hum...mais il peut y avoir des chemins plus courts}and then UkRP is decidable.
\end{proof}

%One begins to test whether $Bad \cap \pred^*(Safe) = \emptyset$ then never resilient car une fois dans bad j'y reste sans pouvoir aller dans Safe.\\

If $Bad \cap \pred^*(Safe) \neq \emptyset$, One computes the minimal elements $A$ of $Bad \cap \pred^*(Safe)$.
resilient iff $Bad \subseteq \pred^*(Safe)$  (eq $Bad \cap \pred^*(Safe) = Bad$)

$Bad \cap \pred^*(Safe) \neq Bad$ alors, on peut se limiter à $Bad' = Bad \cap \pred^*(Safe)$ et c'est (Bad',Safe)-resilient.

@@

\subsection{Case: $Safe=\uparrow Safe$ and $BAD=\downarrow BAD$.}

@@

%Transfering the abstract resilience problems into this framework,
%it is therefore reasonable to demand that both propositions, Safe and BAD, are given by 
%upward-closed or downward-closed sets.

@@

\begin{proposition}\label{down-up}
Let $\mathscr{S}=(S,\rightarrow, \leq)$ be a completion-post-effective $\omega^2$-WSTS with strong compatibility and two finite sets: $I_{BAD}$ and $Min_{Safe}$.
The uniform resilience problem (URP) and the uniform bounded resilience problem (UBRP) are decidable.
The uniform $k$-resilience problem (UkRP) is decidable if moreover $\mathscr{S}$ has the predbasis hypothesis.
\end{proposition}

\begin{proof}
Let $I_{BAD}=\{J_1, J_2,...,J_n\}$ be the ideal decomposition of $BAD$ and $B_{Safe}=\{b_1,b_2,...,b_m\}$ be the minimal basis of $Safe$.
The uniform resilience problem (URP) is equivalent to an infinite number of instances of the coverability problem : for all $x \in BAD$ does there exist an $j$ such that $b_j$ is coverable from $x$. This infinite set of coverability questions can be reduced to $n$ instances of the coverability problem in the completion $\hat{\mathscr{S}}=(Ideals(S),\rightarrow, \subseteq)$ of $\mathscr{S}=(S,\rightarrow, \leq)$.

@@

To decide the UkRP, we may compare the input $k$ with $K$. If $k \geq K$, we conclude that is uniformely $k$-resilient; if $k < K$ we cannot answer. if moreover $\mathscr{S}=(S,\rightarrow, \leq)$ has the predbasis hypothesis, we may compute the least $n$ such that $ \uparrow \pred^n(Safe)=  \uparrow \pred^*(Safe)$. S is uniformely $k$-resilient iff $n \leq k$ \alain{hum...mais il peut y avoir des chemins plus courts}and then UkRP is decidable.
\end{proof}

@@

\subsection{Case: $Safe=\uparrow Safe$ and $BAD=\uparrow BAD$.}

@@

\subsection{Case: $Safe=\downarrow Safe$ and $BAD=\downarrow BAD$.}

%%%%%%%%
\subsection{Case: $Safe=\downarrow Safe$ and $BAD=\uparrow BAD$.}
%
%		cas exclusion mutuelle
%
je prends les minimaux de bad et les idéaux de Safe, pas sur que ce soit decidable....$pred(Safe)$

\begin{tabular}{ l c r }
   BAD/Safe & Up & Down \\
   Up & dec ? & dec ? \\
   Down & dec & ? \\
 \end{tabular}

@@

Uniform resilience is a strong property since it implies that from bad there must exist a path to Safe. But, when one considers a system with an initial state $s_0$, it could be sufficient to ask that only from $post^*(s_0) \cap Bad$, there must exist a path to Safe. However, this condition seems more difficult since one also must decide whether a state $s \in Bad$ is reachable from $s_0$.
%
%\problemx{(I,J)-$k$-resilience problem for WSTS}
%{A state $s$ of a WSTS $(S,\rightarrow, \leq)$, an effective set $I$ (with a given basis), a set $J$.}
%{$\forall s' \in J ~ (s \rightarrow^* s') \implies \exists s'' \in I ~ s' \rightarrow^{\leq k} s''$ ?\newline}
%
%\problemx{general $k$-resilience problem for WSTS}
%{A state $s$ of a WSTS $(S,\rightarrow, \leq)$, an effective set $I$ (with a given basis).}
%{$\forall s'  ~ (s \rightarrow^* s') \implies \exists s'' \in I ~ s' \rightarrow^{\leq k} s''$ ?\newline}
